---
title: "Data Analytics Portfolio Showcase"
author: "Gilbert Morgan"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

# Data Analytics Portfolio Showcase

This Quarto document demonstrates reproducible data analysis skills using R, including data cleaning, exploratory analysis, summary statistics, visualization, and interpretation.

## Skills Demonstrated

- R programming
- Quarto reporting
- Data cleaning and preparation
- Exploratory data analysis
- Summary statistics
- Data visualization
- Grouped analysis
- Reproducible research documentation

## Load Packages

```{r}
library(tidyverse)
```

## Create Sample Dataset

```{r}
analytics_data <- tibble(
  project_area = c(
    "Healthcare Analytics",
    "Machine Learning",
    "SQL Databases",
    "Natural Language Processing",
    "Data Visualization",
    "Statistical Modeling"
  ),
  projects_completed = c(4, 5, 6, 4, 7, 5),
  skill_score = c(88, 91, 86, 84, 90, 89)
)

analytics_data
```

## Summary Statistics

```{r}
analytics_data |>
  summarise(
    total_projects = sum(projects_completed),
    average_skill_score = mean(skill_score),
    highest_skill_score = max(skill_score),
    lowest_skill_score = min(skill_score)
  )
```

## Project Area Ranking

```{r}
analytics_data |>
  arrange(desc(skill_score))
```

## Data Visualization

```{r}
ggplot(analytics_data, aes(x = reorder(project_area, skill_score), y = skill_score)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Analytics Skill Areas",
    x = "Project Area",
    y = "Skill Score"
  ) +
  theme_minimal()
```

## Interpretation

This report highlights applied data analytics skills across multiple technical areas. The workflow demonstrates how R and Quarto can be used to combine code, output, visualization, and written interpretation in one reproducible document.

The analysis reflects a professional reporting workflow that can be expanded for coursework, portfolio projects, healthcare analytics, SQL-based data projects, machine learning documentation, or statistical analysis.
